A Multiresolution Framework for One Dimensional Neural Field Modelling

### 2011-02-27 Multiresolution_framework/ltx/manuscript.tex
edit to  EM algorithm

where the local post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$, parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$ where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The spatial connectivity between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$, for an isotropic and homogeneous neural field. The connectivity kernel is typically a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function $f\left(.\right)$. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, we denote hereafter the current time step by $t$ and the future time step by $t+1$. Here we assume that the sigmoid activation function, $f(\cdot)$, can be approximated by the piecewise function

@@

To derive the state-space representation of the model we substitute decompositions \eqref{eq:KernelFiniteExpansion} and \eqref{eq:FieldFiniteExpansion} into equation \eqref{eq:DiscreteTimeModel} and then cross-multiply the decomposed model by $\boldsymbol\mu\left(\mathbf r\right)$ and integrate over the spatial domain, $\Omega$, giving
\begin{align}\label{eq:DecomposedModel2} 
	&\int_{\Omega} \boldsymbol\mu  \left(\mathbf{r}\right)\boldsymbol\mu\left(\mathbf{r}\right)^\top  d\mathbf{r}\mathbf{x}_{t+1}= 
	\xi\int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\boldsymbol\mu\left(\mathbf{r}\right)^\top d\mathbf{r}\mathbf{x}_t \nonumber \\
	&+T_s \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right) \int_\Omega { 
	   \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r'}\right)
	    \boldsymbol\mu^\top\left(\mathbf{r'}\right)\mathbf{x}_t 
	\, d\mathbf{r}'d\mathbf{r}}
	+ \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r}, 
\end{align}
Now defning 
\begin{equation}\label{eq:Lambdax}
 \mathbf{\Lambda}_{x} \triangleq \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\boldsymbol\mu^\top\left(\mathbf{r}\right) d\mathbf{r}
\end{equation}
\begin{equation}\label{eq:Lambdatheta}
 \mathbf{\Lambda}_{\theta}\triangleq\int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right) \int_\Omega { 
	   \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r'}\right)
	    \boldsymbol\mu^\top\left(\mathbf{r'}\right)\mathbf{x}_t 
	\, d\mathbf{r}'d\mathbf{r}}
\end{equation}
and substituting these into \eqref{eq:DecomposedModel2} and cross-multiplying by $\mathbf{\Lambda}_{x}^{-1}$ yields
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1}=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t
\end{equation}
where
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)=(T_s\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I)
\end{equation}
and $\mathbf I$ is the identity matrix, and 
\begin{equation}\label{eq:Disturbance}
\mathbf w_t= \mathbf{\Lambda}_{x}^{-1}\int_{\Omega}\boldsymbol\mu \left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r}
\end{equation}
which is a vector valued, zero-mean normally distributed white noise process with covariance \cite{Scerri2009}
\begin{equation}
\boldsymbol\Sigma_w =\mathbf{\Lambda}_{x}^{-1}\iint\limits_{\boldsymbol\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\eta\left(\mathbf{r}-\mathbf{r'}\right)\boldsymbol\mu^{\top}\left(\mathbf{r'}\right)d\mathbf{r'}d\mathbf{r}\mathbf{\Lambda}_{x}^{-\top}.
\end{equation}
Similarly, the observation equation of the state-space model is found by substituting decomposition \eqref{eq:FieldFiniteExpansion}
 into equation \eqref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{equation}
where each element of the observation matrix is 
\begin{equation}
	\mathbf{C}_{ij} \triangleq \int_{\Omega}m(\mathbf{r}_i - \mathbf{r}')\boldsymbol{\mu}_j(\mathbf{r}') \, d\mathbf{r}'.
\end{equation}
This completes the state-space representation of the multiresolution approximation of the neural field model. 
% \parham{Mike you can add state-space derivation using Galerkin projection here.}
\section{B-spline wavelet IDE model}
Multiresolution approximation can be implemented using different classes of scaling and wavelet functions. In order to determine the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$ convolution and inner product expressions of the basis functions need to be computable. B-spline wavelet and scaling functions are particularly convenient in the above decomposition as their convolution and inner product expressions can be found analytically.

@@

% Where the subscript $m$ is introduced to $\phi_{j,l}(s)$ and $\psi_{j,l}(s) $ to indicate the $m$th order B-spline scaling and wavelet functions. In the two dimensional case, the convolution integral can be factorised as follows, e.g. 

@@

 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=2T_s\boldsymbol\upsilon_0-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta-2T_s\boldsymbol\upsilon_1

@@

@@

@@

and
 \boldsymbol\Upsilon=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}.

@@

The $\left(i,j\right)^{th}$ element of the  block matrix $\mathbf U_{n_x \times n_x}$ is given by

@@

where $[.]_{j:} $ denotes $j^{th}$ row of the matrix.
% The complexity of this algorithm is $O\left(n_{x}^2n_{\theta}^2\right)+O \left(Tn_{x}^{3}\right)+O\left(Tn_{y}^3\right) $ showing that

>>>>

where the local post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$, parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$ where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The spatial connectivity between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$, for an isotropic and homogeneous neural field. The connectivity kernel is typically a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function $f\left(.\right)$. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, we denote hereafter the current time step by $t$ and the future time step by $t+1$. \parham{Dean, can you replace it by linear firing rate plus referencing plz?}Here we assume that the sigmoid activation function, $f(\cdot)$, can be approximated by the piecewise function

@@

To derive the state-space representation of the model we substitute decompositions \eqref{eq:KernelFiniteExpansion} and \eqref{eq:FieldFiniteExpansion} into equation \eqref{eq:DiscreteTimeModel} and then cross-multiply the decomposed model by $\boldsymbol\mu\left(\mathbf r\right)$ and integrate over the spatial domain, $\Omega$, giving
\begin{align}\label{eq:DecomposedModel2} 
	&\int_{\Omega} \boldsymbol\mu  \left(\mathbf{r}\right)\boldsymbol\mu\left(\mathbf{r}\right)^\top  d\mathbf{r}\mathbf{x}_{t+1}= 
	\xi\int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\boldsymbol\mu\left(\mathbf{r}\right)^\top d\mathbf{r}\mathbf{x}_t \nonumber \\
	&+T_s \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right) \int_\Omega { 
	   \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r'}\right)
	    \boldsymbol\mu^\top\left(\mathbf{r'}\right)\mathbf{x}_t 
	\, d\mathbf{r}'d\mathbf{r}}
	+ \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r}, 
\end{align}
Now defning 
\begin{equation}\label{eq:Lambdax}
 \mathbf{\Lambda}_{x} \triangleq \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\boldsymbol\mu^\top\left(\mathbf{r}\right) d\mathbf{r}
\end{equation}
\begin{equation}\label{eq:Lambdatheta}
 \mathbf{\Lambda}_{\theta}\triangleq\int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right) \int_\Omega { 
	   \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r'}\right)
	    \boldsymbol\mu^\top\left(\mathbf{r'}\right)\mathbf{x}_t 
	\, d\mathbf{r}'d\mathbf{r}}
\end{equation}
and substituting these into \eqref{eq:DecomposedModel2} and cross-multiplying by $\mathbf{\Lambda}_{x}^{-1}$ yields
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1}=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t
\end{equation}
where
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)=(T_s\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I)
\end{equation}
and where $\mathbf I$ is the identity matrix, and 
\begin{equation}\label{eq:Disturbance}
\mathbf w_t= \mathbf{\Lambda}_{x}^{-1}\int_{\Omega}\boldsymbol\mu \left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r}
\end{equation}
which is a vector valued, zero-mean normally distributed white noise process with covariance \cite{Scerri2009}
\begin{equation}
\boldsymbol\Sigma_w =\mathbf{\Lambda}_{x}^{-1}\iint\limits_{\boldsymbol\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\eta\left(\mathbf{r}-\mathbf{r'}\right)\boldsymbol\mu^{\top}\left(\mathbf{r'}\right)d\mathbf{r'}d\mathbf{r}\mathbf{\Lambda}_{x}^{-\top}.
\end{equation}
Similarly, the observation equation of the state-space model is found by substituting decomposition \eqref{eq:FieldFiniteExpansion}
 into equation \eqref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{equation}
where each element of the observation matrix, $\mathbf C$ is given by
\begin{equation}
	\mathbf{C}_{ij} \triangleq \int_{\Omega}m(\mathbf{r}_i - \mathbf{r}')\boldsymbol{\mu}_j(\mathbf{r}') \, d\mathbf{r}'.
\end{equation}
This completes the state-space representation of the multiresolution approximation of the neural field model. 
% \parham{Mike you can add state-space derivation using Galerkin projection here.}
\section{B-spline wavelet IDE model}
Multiresolution approximation can be implemented using different classes of scaling and wavelet functions. In order to determine the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$ convolution and inner product expressions of the basis functions need to be computable. B-spline wavelet and scaling functions are particularly convenient in the above decomposition as their convolution and inner product expressions can be found analytically.

@@

 Where the subscript $m$ is introduced to $\phi_{j,l}(s)$ and $\psi_{j,l}(s) $ to indicate the $m$th order B-spline scaling and wavelet functions.  
% In the two dimensional case, the convolution integral can be factorised as follows, e.g. 

@@

 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta

@@

\begin{equation}\label{eq:upsilon0}

@@

\begin{equation}\label{eq:upsilon1}

@@

\begin{equation}\label{eq:Upsilon}
 \boldsymbol\Upsilon=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i},

@@

and $\beta$ is constant with respect to the connectivity kernel parameters. The $\left(i,j\right)^{th}$ element of the  block matrix $\mathbf U_{n_x \times n_x}$ is given by

@@

where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in equations (\ref{eq:upsilon0}-\ref{eq:Upsilon}) can be computed before EM implementation which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables can be calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$ and the cross-covariance matrix $\mathbf M_t$ using \cite{Shumway2000}
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)
